\plaintitle

### 2025-06-02 docs/style/article-body.tex
added ent, conclusion

\hypertarget{evaluation}{%
\subsubsection{5.3 Evaluation}\label{evaluation}}

To evaluate the correctness of each implementation, a baseline was generated as discussed in section 5.2. Using the naive, initial implementation to process bits and saving for later evaluation gave us a source of truth against which to compare following iterations. To verify that the algorithm successfully removed the patterns it should, we verified the measured entropy score with the command line utility \texttt{ent}\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}.

@@

\begin{table}[H] \centring
\begin{tabularx}{\columnwidth}{|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|}
\hline \textbf{Data structure} & \multicolumn{1}{c|}{\textbf{Teensy ($\mu s$)}}
\\ \hline array & 0.4284 \\ unordered_map & 31.5090 \\ bitset & 0.0474 \\ \hline
\end{tabularx} \caption{Iteration 6 - Data type exloration} \label{tab:iter6}
\end{table}

@@

\hypertarget{conclusion}{%
\subsection{7 CONCLUSION}\label{conclusion}}

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In (3), we calculate the average execution speed required to be \(2.667 \mu\) for the ADC that requires soldering to the microcontroller. From Tables \ref{tab:iter5} and \ref{tab:iter6} we demonstrate execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g. half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667/2 \mu s \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.

@@

\hypertarget{changelog}{%
\subsection{CHANGELOG}\label{changelog}}

2025-02-14: Added background section, smaller reviews to introduction.

2025-02-28: Template adjusted, added methodology. Started review of background and theory to add stronger correlation to computer science. Not yet finished due to review of articles as well as some additional information required from the project owner. The update to theory and background should be considered a heavy work in progress at this stage.

2025-03-10: Moved evaluation down in the methodology in order to provide a better flow. Elaborated further on Toeplitz extraction and ADC converters, as well as motivating the selection of these. Some additional information added in introduction as motivation for the work.

2025-03-13: Elaborated on background, as well as adding more details regarding hardware. Note that the hardware selected is subject to change over time. Further elaborated on related works in optimizing Toeplitz extraction.

2025-04-22: Begun including details regarding initial experimentation, updating details regarding experiments that had to change (e.g.~no baseline on separate hardware).

2025-04-27: Added first table with test data.

2025-05-20: Unfortunately, we've missed adding concrete details about changes made up until this point. Fix later.

>>>>

\subsubsection{5.3 Evaluation}\label{evaluation}

To evaluate the correctness of each implementation, a baseline was generated as discussed in section 5.2. Using the naive, initial implementation to process bits and saving for later evaluation gave us a source of truth against which to compare following iterations. To verify that the algorithm successfully removed the patterns it should, we verified the measured entropy score with the command line utility \texttt{ent}\footnote{\href{https://manpages.ubuntu.com/manpages/trusty/man1/ent.1.html}{Manual page for \texttt{ent}, accessed 2025-04-23.}}. This tool verifies the statistical randomness of any given file, and does several calculations to provide an estimate of how random any given binary file is.

The way \texttt{ent} calculates the entropy of a given binary file is by checking the frequency of certain byte values. If some byte values are more frequent, the binary file has a low entropy -- essentially singifying that there are predictable patterns. Conversely, if all byte values appear roughly equally, the binary file has a high entropy -- meaning that it is more random, and less predictable. The maximum entropy of any given file is \(\log_2(256) = 8 \text{ bits per byte}\), meaning that a perfectly random binary file has an entropy score of 8. For the purposes of our work, we consider values greater than \(7.99\) to be acceptable.

Furthermore, \texttt{ent} also provides a measurement of the possible compression ratio of the file, e.g.~how much more the file can be theoretically compressed. This is calculated as \(1 - \frac{\text{entropy}}{\text{maximum entropy}}\), signifying that the file cannot be compressed further -- that the data is already as compressed as its random distribution allows. Essentially, this informs us that the file has no more detectable statistical patterns or compressable regularities remaining. We consider any file where \texttt{ent} reports a compression ratio of \(0\%\) to be acceptable.

@@

\begin{table}[H] \centring
\begin{tabularx}{\columnwidth}{|>{\centering\arraybackslash}X|>{\centering\arraybackslash}X|}
\hline \textbf{Data structure} & \multicolumn{1}{c|}{\textbf{Teensy ($\mu s$)}}
\\ \hline array & 0.4284 \\ unordered_map & 31.5090 \\ bitset & 0.0474 \\ \hline
\end{tabularx} \caption{Iteration 6 - Data type exploration} \label{tab:iter6}
\end{table}

@@

\subsection{7 DISCUSSION}\label{discussion}

Observing the results, we can place them into the context of the limits imposed by the ADC, as discussed in Section 3.2. In algorithm (3) (\emph{as seen in Section 3.3}), we calculate the average execution speed required to be \(2.667 \mu\) for the ADC that requires soldering to the microcontroller. From Tables \ref{tab:iter5} and \ref{tab:iter6} we demonstrate execution speeds well below that, with iteration 5 as well as iteration 6 -- specifically with bitsets -- being the most prominent results to discuss. Here, our implementation achieved speeds of \(0.0501 \mu s\) and \(0.0474 \mu\) respectively which within the bounds of the ADC listed in Section 3.2. The built in ADC on Teensy has an effective sample rate of 1 MSPS (\emph{e.g.~half of MAX11102AUB}), which would require an average execution speed of \textasciitilde{}\(2.667/2 \mu s \approx 1.334 \mu s\), making our implementation suitable even for that limited conversion speed.

@@

Any requirement of output integers larger than 32 bits may prove problematic. Whereas the bits processed by the algorithm does show an entropy measure quite close to 8, e.g.~reliably random, this does not necessarily mean that two sets of 32 bits can be combined on the host machine to produce a 64-bit integer. Combining these two sets may very well result in a less random result -- in essence, an entropy score that is not entirely random. As quantum computing evolves, the process of bruteforcing algorithms with all possible numbers becomes less of an impossible task, which is why this process of generating random numbers is required. The purpose of using quantum flucuation for random number generation is to provide greater assurances of security not possible from traditional, pseudo-random RNG, and being so severely limited by the size of the output is a major flaw in this implementation worth considering.

\subsection{8 CONCLUSION}\label{conclusion}

This work set out to investigate the feasibility of implementing Toeplitz extraction on resource constrained hardware. Furthermore, we aimed to conclude whether the extraction algorithm could be executed fast enough to not be the limiting factor, e.g.~that the speed of the extraction did not exceed the speed of the ADC conversion. Ostensibly, these investigations were both successful, considering that we managed to create an implementation that executes in roughly one half microsecond on average, far exceeding the limits imposed by the ADC. The optimization with the algorithm revolved around removing as much complexity as possible, avoiding anything that is not the simplest form of data structure for maximum performance gains.

As mentioned in Section 7, this comes at a price. The implementation that did indeed perform above expectation could only output a 32-bit integer, as we cannot take any larger input than 64 bits without any more complex data structure. How much of a flaw this is for a system utilizing this method of OQRNG is beyond the scope of this thesis to investigate. However, it is likely that a larger output is required for any application revolving around tight security. For less critical systems and less essential functionality, this method might, however, prove just secure enough. For instance, an IoT-device that requires a random number output for symmetric cryptography when connecting to a user device may be an ideal candidate for this method of OQRNG.

\subsection{9 FUTURE WORK}\label{future-work}

The main limitation for producing output with a larger size is the lack of a suitable data type in the limited standard library available for MCUs. Furthermore, it appears as if generic containers (\emph{such as \texttt{std::vector}}) that have already been severely optimized still impose a far greater performance penalty for this purpose. Future work may focus on implementing custom data types for larger bit sizes, for instance 128-bit and 256-bit integers. It may be possible that external libraries exist that alredy provide implementations of these integer sizes, but even then a slimmed down and customized implementation may prove the better alternative for this extreme performance constraint.
